---
title: "Exploratory Analysis - Diamonds Tibble"
author: "UB Orengo"
date: "5/6/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIBRARIES USED

```{r}
library(tidyverse)
library(dplyr)
library(magrittr)
library(ggplot2)
library(stats)
library(scales)
```

# DATASET EXPLORED

```{r}
head(diamonds)
```

# EXPLORING X Y Z VALUES

## INTIAL HISTOGRAMS

```{r}
diamonds_hist <- ggplot(diamonds) + 
                    geom_histogram(mapping = aes(x = x), binwidth = 1)
```

### This histogram for the x value has a bit of a left skew

```{r}
diamonds_hist2 <- ggplot(diamonds) + 
                    geom_histogram(mapping = aes(x = y), binwidth = 1)

diamonds_hist3 <- ggplot(diamonds) + 
                    geom_histogram(mapping = aes(x = z), binwidth = 1)
print(diamonds_hist)
print(diamonds_hist2)
print(diamonds_hist3)
```

### We are unable to see the distribution properly of the y and z values because of outlier values

```{r}
print(min(diamonds$y))
print(max(diamonds$y))

print(min(diamonds$z))
print(max(diamonds$z))
```
### After looking at the max/min values and observing the histograms, we create a subset of the current table without the outliers

```{r}
out_diamond_y <- diamonds[which(diamonds[,9]>10),]

out_diamond_z <- diamonds[which(diamonds[,10]>9),]

diamonds_subset_y <- subset(diamonds, diamonds$y <= 11)

diamonds_subset_z <- subset(diamonds, diamonds$z <= 8)
```

## SUBSET GRAPHS -- REMOVED OUTLIERS

### We print new histograms, which are both semi-normally distributed, with new bins

```{r}
diamonds_hist4 <- ggplot(diamonds_subset_y) + 
  geom_histogram(mapping = aes(x = y), binwidth = 1)

diamonds_hist5 <- ggplot(diamonds_subset_z) + 
  geom_histogram(mapping = aes(x = z), binwidth = .5)

print(diamonds_hist4)
print(diamonds_hist5)
```

# EXPLORING PRICE OF DIAMONDS

## Histogram of Distribution of Price of Diamonds

```{r}
diamonds_price <- ggplot(diamonds) +
                    geom_histogram(mapping = aes(x = price), binwidth = 100, color="white")

print(min(diamonds$price))
print(max(diamonds$price))

print(diamonds_price)
```
### It seems that a large portion of the diamonds in the dataset are between the first 10 bins, which would be 0-1000$

# COMPARISION OF .99 CARAT DIAMONDS TO 1 CARAT DIAMONDS

```{r}
carat_count <- diamonds[which(diamonds[,1] == 0.99),]
view(carat_count)
```

### 23 rows in diamonds$carat = 0.99, there are 23 diamonds in the dataset that equal .99 carats

```{r}
carat_count2 <- diamonds[which(diamonds[,1] == 1),]
view(carat_count2)
```

### 1,558 rows in diamonds$carat = 1. there are 1,558 diamonds in the dataset that equal 1 carat. 

### The difference may be attributed to values being rounded up to be able to price the diamonds at 1 carat, as opposed to .99 carats.

## Summary Table -- Average Price by Carat

```{r}
avg_price_by_carat <- diamonds %>%
                        group_by(carat) %>%
                        summarize(avg_price = mean(price))

price_99_carat <- avg_price_by_carat[which(avg_price_by_carat[,1] == .99),]
price_1_carat <- avg_price_by_carat[which(avg_price_by_carat[,1] == 1),]

print(avg_price_by_carat)
print(price_99_carat)
print(price_1_carat)
```

## Line Plot of Average Price by Number of Carats

```{r}
carat_plot <- ggplot(data = avg_price_by_carat, mapping = aes(x = carat, y = avg_price)) +
                  geom_line() +
                  geom_point() +
                  geom_smooth() +
                  ggtitle("Average Price By Carat") +
                  xlab("Number of Carats") + ylab("Average Price")

print(carat_plot) + ggtitle("Average Price By Carat") + xlab("Number of Carats") + ylab("Average Price")
```

### You can see a distinct slope between .99 carat and 1 carat

```{r}
print(price_99_carat)
print(price_1_carat)
price_diff <- price_1_carat[,2]-price_99_carat[,2]                   
print(price_diff)
```

### There is an over 800$ difference between the average price of a .99 carat diamond and a 1 carat diamond
